Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-04-05 paper/manuscript.Rmd
trying to make an intro that makes sense

@@

# introduction
Replicability is a desirable feature for any study that we build theory or future experiments off of. It's a pre-requisite for building theories that we have phenomena for these studies to explain, but results that cannot be observed consistently aren't results to build theories around. However, to be a load-bearing empirical result, a study needs to do much more than just replicate, it also needs to be interesting, have a believable theoretical interpretation, generalize appropriately, and have reliable constructs.

The replication rates in psychology are low, with estimates of replicability of experimental findings hovering somewhere around 50%.   A major concern is that published literature may not replicate as much as previously thought or as desired. This has been illustrated by prominent findings/theories not replicating at all (ego depletion/ TMT? CITATIONS). Some prominent non-replications and concerns about research practices that might impact replication rates led to a few large-scale replication projects, aiming to quantify replication rates across swaths of the literature.

We are aware of three large-scale replication efforts replicating experimental results in the existing psychology literature. RP:P sampled roughly 100 studies from top psychology journals in 2008 [@openscienceconsortium2015]. They found an overall replication rate around 40%, which provided evidence to support growing concerns about the non-reliability of the literature.The ManyLabs series of studies have also done large-scale replications of effects from psychology. Due to their primary goal of investigating different forms of hetereogeneity, their sampling has been non-representive, focusing on short studies with only two conditions. Across Many labs 1-3 foo bar replicated [@klein2014; @klein2018; @ebersole2016]. @camerer2018 replicated the 21 behaviors studies published in Nature and Science from 2010-2015 that did not require special populations or special equipment. They found a roughly 60% replication rate.

These replication studies, and similar ones in other fields, open up big questions about why the replication rates are so low. There are many different theories both about causes and motivations for non-replications (relating to the incentives and research practices of science) and about what studies are most affected by these. There have been both experimental and correlational approaches to understanding the replication rates.

A couple studies have attempted to intervene on the low replication rate, but seeing if certain factors can affect whether studies replicate. For instance, @protzko was a prospective replication study, showing that with a slew of best practices and large samples size (TODO look up details), the replications of FOOBAR novel whatevers had similar effect sizes to the original. Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at TODO CITE. These types of experiments are valuable, but they aren't very scalable because they are expensive and some potential influences on non-replication are not experimentally manipulable. 
Instead, many studies have pursued correlational results about what predicts replication success, whether via regression models, prediction markets and elicitations, or machine learning techniques. However, all of these approaches depend on data from replications. While many replications have been conducted, most were not randomly sampled and are not pooled together, so these studies instead rely on the small number of data points, pulling heavily from RP:P in particular. The RP:P dataset was itself much discussed and reanalyzed [@etz2016;@gilbert2016;@patil2016;@anderson2016].

Our contribution to this literature is to provide a collection of 176 replication projects, done by students for a class. This is a sizeable number of projects, compared to the existing amount of data.

>>>>

 as validity, interestingness, and generalizability, can also be important
Cumulative science and theory-building rely on empirical results that Replicability is a desirable feature for any study that we build theory or future experiments off of. It's a pre-requisite for building theories that we have phenomena for these studies to explain, but results that cannot be observed consistently aren't results to build theories around. However, to be a load-bearing empirical result, a study needs to do much more than just replicate, it also needs to be interesting, have a believable theoretical interpretation, generalize appropriately, and have reliable constructs.

@@

# introduction
[wacky metaphor, maybe drop] When building a stone wall, it's important that the components be stones that will retain their shape and not clots of dirt and sand that might fall apart and bring the rest of the wall falling down. Building a cumulative body of scientific theories similarly requires that the empirical results, especially the more foundational empirical results, be sturdy. Results that cannot be observed consistently are like dirt clots, they don't hold their shape and thus are not a strong foundation for the theories and studies built on top of them. Replications and the theory of replicability can be thought of as a way of testing whether the empirical results are sturdy. Replicating is a necessary feature for load-bearing empirical results. We note that replicability is not the only desirable feature of a study, and that replicating does not mean that a study is valid, or generalizable, or interesting.

A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all (CITE ex ego depletion, TMT), and by large-scale replication projects such as RP:P which found a replication rate around 40% for a sample of the psychology literature [@openscienceconsortium2015]. This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

A big open question is why the replication rates in psychology are so low. There are theoretical approaches modelling how questionable research practices (which many psychologists admit to) and a bias towards positive results would lead to a high rate of (non-replicable) false positives in the literature (CITE). There's some disconnect between these theoretically causal factors and empirical results about what causes or correlates with replicability.

A couple studies have attempted to intervene on the low replication rate, but seeing if certain factors can affect whether studies replicate. For instance, across 16 studies, @protzko2020 showed that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates  that matched theoretical expectations based on the effect sizes and sample sizes, and replication effect sizes comparable with the original.  Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at [@ebersole2020]. These types of experiments are valuable, but they aren't very scalable because they are expensive and some potential influences on non-replication are not experimentally manipulable.

As an alternative, much of the work looking at predictors of replicability has been correlation: whether using regression models, prediction markets, or machine learning techniques to build predictions about what will replicate. However, all of these approaches depend on data from replications. While many replications have been conducted, most were not randomly sampled and are not pooled together, so these studies instead rely on the small number of data points from large-scale replication projects. The RP:P dataset itself is much discussed and reanalyzed [@etz2016;@gilbert2016;@patil2016;@anderson2016].

This reliance on the same small set of data is problematic, as much of what we think we know about replicability is overfit to this small sample. However, given the time-consuming nature of running a large number of replications, there is a limited amount of large-scale replication data to draw on. We are aware of three large-scale replication efforts replicating experimental results in the existing psychology literature. RP:P sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40% [@openscienceconsortium2015]. The ManyLabs series of replications investigating heterogeneity has sampled target studies that are short and compare only two conditions, which is not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57%) were found to replicate [@klein2014; @klein2018; @ebersole2016]. @camerer2018 found a roughly 60% replication rate when they replicated all 21 behaviors studies published in Nature and Science from 2010-2015 that did not require special populations or special equipment. Those are all the large-scale replication projects of experimental psychology results available at this time. These roughly 170 results are heavily represented in the work predicting what effects will replicate.

Our main contribution is to add a collection of 176 student replication projects to the available pool. We use these studies to examine the correlates of replicability.

### 2023-04-05 paper/manuscript.tex
trying to make an intro that makes sense

@@

Gilbert et al. (\protect\hyperlink{ref-gilbert2016}{2016}) Anderson et al. (\protect\hyperlink{ref-anderson2016}{2016}) people argue a lot about how to interpret RPP

@@

replication is important (but not everything) - good starting place
* prereq of theory building that the results of interest ``can be observed consistently''
* but not everything (can still have other problems)

The replication rates in psychology are low, and
replication rates are low in psychology (see studies)

big open question: why

experimental and correlational approaches to answering it

experimental is hard - protzko, ML5 etc. but they are expensive and you can't do them for many research questions

correlational has been pursued extensively, but is overfit to small data (prediction markets / ML models / regression of all types)

@@

Psychology is in the middle of heated debate over its research practices and the subject of many reform efforts. Many of these issues concern how trustworth the published literature is (or isn't), how it got to be this way, and what measures can be taken to improve the quality of findings for the future. These issues center around issues of reproducibility and replicability. Concerns about replication rates have become a large point of discussion in psychology and other fields, with large scale replications spurring these discussions and providing the empirical data analyzed by all sides.

A major concern is that published literature may not replicate as much as previously thought or as desired. This has been illustrated by prominent findings/theories not replicating at all (ego depletion/ TMT?). To understand how replicable the literature as a whole is, we need empirical data about replication attempts across a wide range of studies. Most replications are targeted replications of individual results of interest to the researcher, but these are not sampled randomly from the literature, are performed in inconsistent ways, and may be selectively published depending on the results. Thus, they are not ideal for estimating the overall replicability of the field.

There have been a limited number of large-scale replication efforts, which sample studies from a discipline to estimate overall replicability. Due to the arduous process of replicating many studies, there have not been many large-scale replication efforts. Thus, all of the argumentation around predictors of replications and field-wide replication rates is fit to a small number of data points.

\hypertarget{prior-literature}{%
\subsection{prior literature}\label{prior-literature}}

We are aware of three large-scale replication efforts replicating experimental results in the existing psychology literature. RP:P sampled roughly 100 studies from top psychology journals in 2008 (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). They found an overall replication rate around 40\%, which provided evidence to support growing concerns about the non-reliability of the literature. The RP:P dataset was itself much discussed and reanalyzed (\protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) {[}idk, maybe worth framing that this was a big deal study when it came out{]}. TODO some discussion of how many papers have reanalysed this cutting the data different ways.

The ManyLabs series of studies have also done large-scale replications of effects from psychology. Due to their primary goal of investigating different forms of hetereogeneity, their sampling has been non-representive, focusing on short studies with only two conditions. Across Many labs 1-3 foo bar replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Many labs 5 was a re-replication attempt on 10 of RP:P and rescued 2/10 (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). TODO look up details

Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated the 21 behaviors studies published in Nature and Science from 2010-2015 that did not require special populations or special equipment. They found a roughly 60\% replication rate.

@@

can even cite the ``science is not a signal detection problem'' paper on not randomly sampling from the literature.

Prior approaches to replicability have focused on a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (van bavel etc.) or even time period (ramscar).

@@

In contrast, we have pursued a different estimand: the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they feel that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation. SPELL OUT SAMPLING DIFFERENCE AND SUCCESS DIFFERENCE.

>>>>

To understand how replicable the literature as a whole is, we need empirical data about replication attempts across a wide range of studies. Most replications are targeted replications of individual results of interest to the researcher, but these are not sampled randomly from the literature, are performed in inconsistent ways, and may be selectively published depending on the results. Thus, they are not ideal for estimating the overall replicability of the field.

There have been a limited number of large-scale replication efforts, which sample studies from a discipline to estimate overall replicability. Due to the arduous process of replicating many studies, there have not been many large-scale replication efforts. Thus, all of the argumentation around predictors of replications and field-wide replication rates is fit to a small number of data points.
Psychology is in the middle of heated debate over its research practices and the subject of many reform efforts. Many of these issues concern how trustworth the published literature is (or isn't), how it got to be this way, and what measures can be taken to improve the quality of findings for the future. These issues center around issues of reproducibility and replicability. Concerns about replication rates have become a large point of discussion in psychology and other fields, with large scale replications spurring these discussions and providing the empirical data analyzed by all sides.

{[}idk, maybe worth framing that this was a big deal study when it came out{]}. TODO some discussion of how many papers have reanalysed this cutting the data different ways.

Many labs 5 was a re-replication attempt on 10 of RP:P and rescued 2/10 (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). TODO look up details

@@

@@

The replication rates in psychology are low, with estimates of replicability of experimental findings hovering somewhere around 50\%. A major concern is that published literature may not replicate as much as previously thought or as desired. This has been illustrated by prominent findings/theories not replicating at all (ego depletion/ TMT? CITATIONS). Some prominent non-replications and concerns about research practices that might impact replication rates led to a few large-scale replication projects, aiming to quantify replication rates across swaths of the literature.

We are aware of three large-scale replication efforts replicating experimental results in the existing psychology literature. RP:P sampled roughly 100 studies from top psychology journals in 2008 (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). They found an overall replication rate around 40\%, which provided evidence to support growing concerns about the non-reliability of the literature.The ManyLabs series of studies have also done large-scale replications of effects from psychology. Due to their primary goal of investigating different forms of hetereogeneity, their sampling has been non-representive, focusing on short studies with only two conditions. Across Many labs 1-3 foo bar replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated the 21 behaviors studies published in Nature and Science from 2010-2015 that did not require special populations or special equipment. They found a roughly 60\% replication rate.

These replication studies, and similar ones in other fields, open up big questions about why the replication rates are so low. There are many different theories both about causes and motivations for non-replications (relating to the incentives and research practices of science) and about what studies are most affected by these. There have been both experimental and correlational approaches to understanding the replication rates.

A couple studies have attempted to intervene on the low replication rate, but seeing if certain factors can affect whether studies replicate. For instance, (\protect\hyperlink{ref-protzko}{\textbf{protzko?}}) was a prospective replication study, showing that with a slew of best practices and large samples size (TODO look up details), the replications of FOOBAR novel whatevers had similar effect sizes to the original. Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at TODO CITE. These types of experiments are valuable, but they aren't very scalable because they are expensive and some potential influences on non-replication are not experimentally manipulable.
Instead, many studies have pursued correlational results about what predicts replication success, whether via regression models, prediction markets and elicitations, or machine learning techniques. However, all of these approaches depend on data from replications. While many replications have been conducted, most were not randomly sampled and are not pooled together, so these studies instead rely on the small number of data points, pulling heavily from RP:P in particular. The RP:P dataset was itself much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}).

Our contribution to this literature is to provide a collection of 176 replication projects, done by students for a class. This is a sizeable number of projects, compared to the existing amount of data.

@@

@@

Prior approaches to replicability have focused on a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (CITE van bavel etc.) or even time period (CITE ramscar).

@@

In contrast, we have pursued a different estimand: the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they feel that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation.

There's different approaches one could take to sampling, we take the versimilitude approach. Our sample is not a random sample of the literature, or of a given journal, or some weighting, but it does represent studies that people are interested in replicating.

Quantifying when an effect is a replication is hard, and many statistical metrics have been proposed, often with tradeoffs (TODO citations). We instead focus on a subjective score representing a more functionally oriented outcome: studies replicate if the results match the original enough that researchers would feel comfortable building on the study. This matches the motivation for many replications (often in replicate and extend paradigms). SPELL OUT SAMPLING DIFFERENCE AND SUCCESS DIFFERENCE.

There's a ``can I trust this finding'' question, but a key part of it is, ``if I did this, would I get the same pattern of results'' -- and that's what we're measuring.
